Gammapy: A Python package for gamma-ray astronomy

### 2023-01-20 src/ms-review.tex
Adjust code example font size
Remove fit section todo

% Note: not sure if we want to use that ... doesn't look too pretty
\newcommand{\astropy}{Astropy\xspace}
\newcommand{\gammapy}{Gammapy\xspace}
\newcommand{\scipy}{Scipy\xspace}
\newcommand{\numpy}{Numpy\xspace}

\newcommand{\hess}{H.E.S.S.\xspace}
\newcommand{\hawc}{HAWC\xspace}
\newcommand{\veritas}{VERITAS\xspace}
\newcommand{\magic}{MAGIC\xspace}
\newcommand{\astri}{ASTRI\xspace}
\newcommand{\iact}{IACT\xspace}
\newcommand{\iacts}{IACTs\xspace}
\newcommand{\cta}{CTA\xspace}
\newcommand{\swgo}{SWGO\xspace}
\newcommand{\irf}{IRF\xspace}
\newcommand{\irfs}{IRFs\xspace}
\newcommand{\fermi}{\textit{Fermi}-LAT\xspace}
\newcommand{\gammaray}{$\gamma$-ray\xspace}
\newcommand{\gammarays}{$\gamma$-rays\xspace}
\newcommand{\gadf}{GADF\xspace}
\newcommand{\milagro}{MILAGRO\xspace}
\newcommand{\github}{GitHub\xspace}
\newcommand{\iminuit}{IMinuit}

@@

The DL3 level data represented by the \code{Observation} class consist
of two types of elements: first a list of \gammaray events with relevant physical
quantities such as estimated energy, direction and arrival
times, which is represented by the \code{EventList} class. Second a set of
associated IRFs, providing the response of the system, typically
factorised in independent components as described in
Section~\ref{ssec:gammapy-irf}. The separate handling of event lists and IRFs
additionally allows for data from non-IACT \gammaray instruments to be read. For
example, to read \fermi data, the user can read separately their event list
(already compliant with the \gadf specifications) and then find the appropriate
IRF class representing the response functions provided by \fermi, see
example in Section~\ref{ssec:multi-instrument-analysis}.
%
\begin{figure}[ht!]
	\import{code-examples/generated/}{gp_data}
	\caption{
        Using \code{gammapy.data} to access DL3 level data with a \code{DataStore} object.
        Individual observations can be accessed by their unique integer observation id number.
        The actual events and instrument response functions can be accessed
        as attributes on the \code{Observation} object, such as \code{.events}
        or \code{.aeff} for the effective area information.
    }
	\label{fig*:minted:gp_data}
\end{figure}
%

@@

\begin{figure}
	\import{code-examples/generated/}{gp_maps}

@@

\subsection{gammapy.datasets}
\label{ssec:gammapy-datasets}
%
\begin{figure}
	\import{code-examples/generated/}{gp_datasets}
	\caption{
        Using \code{gammapy.datasets} to read existing reduced binned datasets.
        After the different datasets are read from disk they are collected into a
        common \code{Datasets} container. All dataset types have an associated
        name attribute to allow access by name later in the code. The
        environment variable \code{\$GAMMAPY\_DATA} is automtically resolved
        by \gammapy.
    }
	\label{fig*:minted:gp_datasets}
\end{figure}
%
The \code{gammapy.datasets} subpackage contains classes to bundle
together binned data along with associated models and the likelihood, which
provides an interface to the \code{Fit} class (Sec \ref{sssec:fit}) for
modeling and fitting purposes. Depending upon the type of analysis and the
associated statistics, different types of Datasets are supported. \code{MapDataset} is
used for 3D (spectral and morphological) fitting, and a 1D spectral fitting is
done using \code{SpectrumDataset}. While the default statistics for both of these is
\emph{Cash}}, their corresponding on off versions are adapted for the case where the
background is measured from real off counts, and support \emph{WStat} statistics. The
predicted counts are computed by convolution of the models with the associated
IRFs. Fitting of precomputed flux points is enabled through \code{FluxPointsDataset},
using chi2 statistics. Multiple datasets of same or different types can be
bundled together in \code{Datasets} (e.g., Figure \ref{fig*:minted:gp_datasets}),
where the likelihood from each constituent member is added, thus facilitating
joint fitting across different observations, and even different instruments
across different wavelengths. Datasets also provide functionalities for
manipulating reduced data, e.g. stacking, sub-grouping, plotting. Users can
also create their customized datasets for implementing modified likelihood
methods.

\subsection{gammapy.makers}
\label{ssec:gammapy-makers}
%
\begin{figure}
    \import{code-examples/generated/}{gp_makers}
	\caption{
        Using \code{gammapy.makers} to reduce DL3 level data into a
		\code{MapDataset}. All \code{Maker} classes represent 
		a step in the data reduction process. They take
        the configuration on initialisation of the class. They 
		also consitently define \code{.run()} methods, which take 
		a dataset object and optionally an \code{Observation} 
		object. This way \code{Maker} classes can be chained
		to define more complex data reduction "pipelines".
    }
	\label{ig*:minted:gp_makers}
\end{figure}
%
The data reduction step includes all tasks required to process and prepare
\gammaray data from the DL3 level to modeling and fitting. The \code{gammapy.makers} sub-package
contains the various classes and functions required to do so. First, events are
binned and IRFs are interpolated and projected onto the chosen analysis
geometry. The end product of the data reduction process are a set of binned counts,
background exposure, psf and energy dispersion maps at the DL4 level. 
The \code{MapDatasetMaker} and \code{SpectrumDatasetMaker} are
responsible for this task, see Fig~\ref{ig*:minted:gp_makers}.

@@

\begin{figure}
	\import{code-examples/generated/}{gp_stats}
	\caption{
        Using \code{gammapy.stats} to compute statistical quantities
        such as excess, signficance and assymetric errors
        from counts based data. The data is passed on initialisation
        of the \code{WStatCountsStatistic} class. The quantities
        are the computed ON excess of the corresponding class
        attributes such as \code{stat.n\_sig} and \code{stat.sqrt\_ts}.
    }
	\label{codeexample:stats} \end{figure}

@@

\begin{figure}
	\import{code-examples/generated/}{gp_models}
	\caption{Using \code{gammapy.modeling.models} to define a source model with a
    spectral, spatial and temporal component. For convenience the model
    parameters can be defined as strings with attached units. The spatial model
    takes an addtional \code{frame} parameter which allow users to define
    the coordinate frame of the position of the model.
    }
	\label{fig*:minted:gp_models}
\end{figure}

@@

The \code{Fit} class provides methods to fit i.e., optimise parameters and estimate
parameter errors and correlations. It interfaces with a \code{Datasets} object, which
in turn is connected to a \code{Models} object containing the model parameters in its
\code{Parameters} object. Models can be unique for a given dataset, or contribute to
multiple datasets, allowing e.g., to do a joint fit to
multiple IACT datasets, or to a joint IACT and \fermi dataset. Many
examples are given in the tutorials.

\todo{Add fit code example...?}

The Fit class provides a uniform interface to multiple fitting backends:

\begin{itemize}
	\item \iminuit~\citep{iminuit}
	\item \code{scipy.optimize}~\citep{2020SciPy-NMeth}
	\item Sherpa~\citep{sherpa-2011}
\end{itemize}

Note that, for now, covariance matrixand errors are computed only for the fitting with 
\iminuit. However depending on
the problem other optimizers can better perform, so sometimes it can be useful
to run a pre-fit with alternative optimization methods. In future we plan to
extend the supported fitting backends, including for example solutions based on Markov chain Monte Carlo.
\footnote{a prototype is available in gammapy-recipes,
	\url{https://gammapy.github.io/gammapy-recipes/_build/html/notebooks/mcmc-sampling-emcee/mcmc_sampling.html}
}

@@

\begin{figure}
	\import{code-examples/generated/}{gp_estimators}
	\caption{Using the \code{TSMapEstimator} object from \code{gammapy.estimators} to compute a
        a flux, flux upper limits and TS map. The additional parameters \code{n\_sigma}
        and \code{n\_sigma\_ul} define the confidence levels (in multiples of the normal distribution width)
        of the flux error and and flux upper limit maps respectively.
    }
    \label{fig*:minted:gp_estimators}
\end{figure}

@@

\begin{figure}
	\import{code-examples/generated/}{gp_catalogs}
	\caption{Using \code{gammapy.catalogs} to access the underlying model, flux points and
		light-curve from the \fermi 4FGL catalog for the blazar PKS 2155-304}
	\label{codeexample:data}
\end{figure}

>>>>

% Note: not sure if we want to use that ... doesn't look too pretty
\newcommand{\astropy}{Astropy\xspace}
\newcommand{\gammapy}{Gammapy\xspace}
\newcommand{\scipy}{Scipy\xspace}
\newcommand{\numpy}{Numpy\xspace}
\newcommand{\iminuit}{IMinuit\xspace}
\newcommand{\sherpa}{Sherpa\xspace}

\newcommand{\hess}{H.E.S.S.\xspace}
\newcommand{\hawc}{HAWC\xspace}
\newcommand{\veritas}{VERITAS\xspace}
\newcommand{\magic}{MAGIC\xspace}
\newcommand{\astri}{ASTRI\xspace}
\newcommand{\iact}{IACT\xspace}
\newcommand{\iacts}{IACTs\xspace}
\newcommand{\cta}{CTA\xspace}
\newcommand{\swgo}{SWGO\xspace}
\newcommand{\irf}{IRF\xspace}
\newcommand{\irfs}{IRFs\xspace}
\newcommand{\fermi}{\textit{Fermi}-LAT\xspace}
\newcommand{\gammaray}{$\gamma$-ray\xspace}
\newcommand{\gammarays}{$\gamma$-rays\xspace}
\newcommand{\gadf}{GADF\xspace}
\newcommand{\milagro}{MILAGRO\xspace}
\newcommand{\github}{GitHub\xspace}

@@

The DL3 level data represented by the \code{Observation} class consist
of two types of elements: first a list of \gammaray events with relevant physical
quantities such as estimated energy, direction and arrival
times, which is represented by the \code{EventList} class. Second a set of
associated IRFs, providing the response of the system, typically
factorised in independent components as described in
Section~\ref{ssec:gammapy-irf}. The separate handling of event lists and IRFs
additionally allows for data from non-IACT \gammaray instruments to be read. For
example, to read \fermi data, the user can read separately their event list
(already compliant with the \gadf specifications) and then find the appropriate
IRF class representing the response functions provided by \fermi, see
example in Section~\ref{ssec:multi-instrument-analysis}.
%
\begin{figure}
	\small
	\import{code-examples/generated/}{gp_data}
	\caption{
        Using \code{gammapy.data} to access DL3 level data with a \code{DataStore} object.
        Individual observations can be accessed by their unique integer observation id number.
        The actual events and instrument response functions can be accessed
        as attributes on the \code{Observation} object, such as \code{.events}
        or \code{.aeff} for the effective area information.
    }
	\label{fig*:minted:gp_data}
\end{figure}
%

@@

\begin{figure}
	\small
	\import{code-examples/generated/}{gp_maps}

@@

\subsection{gammapy.datasets}
\label{ssec:gammapy-datasets}
%
\begin{figure}
	\small
	\import{code-examples/generated/}{gp_datasets}
	\caption{
        Using \code{gammapy.datasets} to read existing reduced binned datasets.
        After the different datasets are read from disk they are collected into a
        common \code{Datasets} container. All dataset types have an associated
        name attribute to allow access by name later in the code. The
        environment variable \code{\$GAMMAPY\_DATA} is automtically resolved
        by \gammapy.
    }
	\label{fig*:minted:gp_datasets}
\end{figure}
%
The \code{gammapy.datasets} subpackage contains classes to bundle
together binned data along with associated models and the likelihood, which
provides an interface to the \code{Fit} class (Sec \ref{sssec:fit}) for
modeling and fitting purposes. Depending upon the type of analysis and the
associated statistics, different types of Datasets are supported. \code{MapDataset} is
used for 3D (spectral and morphological) fitting, and a 1D spectral fitting is
done using \code{SpectrumDataset}. While the default statistics for both of these is
\emph{Cash}}, their corresponding on off versions are adapted for the case where the
background is measured from real off counts, and support \emph{WStat} statistics. The
predicted counts are computed by convolution of the models with the associated
IRFs. Fitting of precomputed flux points is enabled through \code{FluxPointsDataset},
using chi2 statistics. Multiple datasets of same or different types can be
bundled together in \code{Datasets} (e.g., Figure \ref{fig*:minted:gp_datasets}),
where the likelihood from each constituent member is added, thus facilitating
joint fitting across different observations, and even different instruments
across different wavelengths. Datasets also provide functionalities for
manipulating reduced data, e.g. stacking, sub-grouping, plotting. Users can
also create their customized datasets for implementing modified likelihood
methods.

\subsection{gammapy.makers}
\label{ssec:gammapy-makers}
%
\begin{figure}
	\small
    \import{code-examples/generated/}{gp_makers}
	\caption{
        Using \code{gammapy.makers} to reduce DL3 level data into a
		\code{MapDataset}. All \code{Maker} classes represent 
		a step in the data reduction process. They take
        the configuration on initialisation of the class. They 
		also consitently define \code{.run()} methods, which take 
		a dataset object and optionally an \code{Observation} 
		object. This way \code{Maker} classes can be chained
		to define more complex data reduction "pipelines".
    }
	\label{ig*:minted:gp_makers}
\end{figure}
%
The data reduction step includes all tasks required to process and prepare
\gammaray data from the DL3 level to modeling and fitting. The \code{gammapy.makers} sub-package
contains the various classes and functions required to do so. First, events are
binned and IRFs are interpolated and projected onto the chosen analysis
geometry. The end product of the data reduction process are a set of binned counts,
background exposure, psf and energy dispersion maps at the DL4 level. 
The \code{MapDatasetMaker} and \code{SpectrumDatasetMaker} are
responsible for this task, see Fig~\ref{ig*:minted:gp_makers}.

@@

\begin{figure}
	\small
	\import{code-examples/generated/}{gp_stats}
	\caption{
        Using \code{gammapy.stats} to compute statistical quantities
        such as excess, signficance and assymetric errors
        from counts based data. The data is passed on initialisation
        of the \code{WStatCountsStatistic} class. The quantities
        are the computed ON excess of the corresponding class
        attributes such as \code{stat.n\_sig} and \code{stat.sqrt\_ts}.
    }
	\label{codeexample:stats} \end{figure}

@@

\begin{figure}
	\small
	\import{code-examples/generated/}{gp_models}
	\caption{Using \code{gammapy.modeling.models} to define a source model with a
    spectral, spatial and temporal component. For convenience the model
    parameters can be defined as strings with attached units. The spatial model
    takes an addtional \code{frame} parameter which allow users to define
    the coordinate frame of the position of the model.
    }
	\label{fig*:minted:gp_models}
\end{figure}

@@

The \code{Fit} class provides methods to fit i.e., optimise model parameters and estimate
parameter errors and correlations. It interfaces with a \code{Datasets} object, which
in turn is connected to a \code{Models} object containing the model parameters in its
\code{Parameters} object. Models can be unique for a given dataset, or contribute to
multiple datasets, allowing e.g., to do a joint fit to
multiple IACT datasets, or to a joint IACT and \fermi dataset. Many
examples are given in the tutorials.

The \code{Fit} class provides a uniform interface to multiple fitting backends:

\begin{itemize}
	\item \iminuit~\citep{iminuit}
	\item \code{scipy.optimize}~\citep{2020SciPy-NMeth}
	\item \sherpa~\citep{sherpa-2011}
\end{itemize}

Note that, for now, covariance matrix and errors are computed only for the fitting with 
\iminuit. However depending on
the problem other optimizers can better perform, so sometimes it can be useful
to run a pre-fit with alternative optimization methods. In future we plan to
extend the supported fitting backends, including for example solutions based on Markov chain Monte Carlo methods.
\footnote{a prototype is available in gammapy-recipes,
	\url{https://gammapy.github.io/gammapy-recipes/_build/html/notebooks/mcmc-sampling-emcee/mcmc_sampling.html}
}

@@

\begin{figure}
	\small
	\import{code-examples/generated/}{gp_estimators}
	\caption{Using the \code{TSMapEstimator} object from \code{gammapy.estimators} to compute a
        a flux, flux upper limits and TS map. The additional parameters \code{n\_sigma}
        and \code{n\_sigma\_ul} define the confidence levels (in multiples of the normal distribution width)
        of the flux error and and flux upper limit maps respectively.
    }
    \label{fig*:minted:gp_estimators}
\end{figure}

@@

\begin{figure}
	\small
	\import{code-examples/generated/}{gp_catalogs}
	\caption{Using \code{gammapy.catalogs} to access the underlying model, flux points and
		light-curve from the \fermi 4FGL catalog for the blazar PKS 2155-304}
	\label{codeexample:data}
\end{figure}
